---
title: Trace elements accrued in endangered freshwater bivalves in two North Carolina streams
author:
- Bradley C. Saul
- Nathan R. Miller
- Heather Gunn
- Michael Walter
- Scott Salger
- Mike Gangloff
- Jay F. Levine
bibliography: elktoe.bib
toc: false
output: 
  bookdown::pdf_document2:
    number_sections: true
  bookdown::word_document2: default
header-includes:   
- \usepackage{amsmath, setspace}
- \usepackage{multirow}
- \graphicspath{ {figures/} } 
abstract: >
  \noindent Precipitous declines in the Federally endangered Appalachian Elktoe (*Alasmidonta raveneliana*) have been observed since 2004 in the Little Tennessee River, but populations appear stable in the Tuckasegee River. To assess river-specific factors associated with health and physiological outcomes, we conducted a completely randomized experiment, assigning specimens to sites within each river. Here, we examine differences between sites in trace element concentrations accumulated in specimens’ valves during the experiment.
keywords: TODO
---

\doublespacing

# Introduction 

After major hurricane-related flooding in 2004, the population of Federally endangered Appalachian Elktoe (*Alasmidonta raveneliana*) in North Carolina's Little Tennessee River declined precipitously, but within the same watershed's Tuckasegee River populations appear stable. Reasons for the decline are unclear [@jarvis2011water; @miller2013concentrations; @pandolfi2016effects]. To determine how river-specific factors may be associated with health and physiological outcomes, we conducted a completely randomized experiment by assigning eighty-two *A. raveneliana* along with an equal number of hatchery-raised Wavy-Rayed Lampmussel (*Lampsilis fasciola*) to sites within each river.

Similar to tree rings, bivalve shells archive some aspects of their environmental history and have been used as bioindicators of pollution [@markich2002freshwater; @brown2005freshwater; @shoults2014resolution]. Exposure to certain heavy metals have been shown to increase morbidity and mortality in freshwater bivalves [@naimo1995a-review]. @miller2013concentrations evaluated sediment concentrations of Cu, Cr, Ni, and Zn in the Little Tennessee and Tuckasegee. When considered with the chemical concentrations of *A. raveneliana* shells measured by @jarvis2011water, they found these elements (Cu in particular) unlikely to be the primary culprit in the *A. raveneliana* decline.

In this paper, we ask whether and to what degree the exposure to the experimental sites modified the accretion of twenty trace elements in specimens’ valves. We demonstrate how chemical trajectories obtained by laser ablation inductively coupled plasma mass spectrometry (LA-ICP-MS) transects [@pozebon2014review] can be aligned with anatomical features of the valves, thus making comparisons of specific valve layers and time periods possible. We hypothesized that trajectories in the nacral growth that accumulated during the experiment for some elements would be associated with the experimental sites. The assignment to sites was completely randomized, hence we use the randomization distribution [@coppock2019ri2] to draw inference from generalized additive mixed models [@wood2017gam] of the elemental trajectories. 

TODO: one more paragraph with contributions of this paper

The paper is organized as follows. Section \ref{materials-and-methods} outlines the experimental design, data collection, anatomy/chemical alignment procedures, and the statistical methods. Results are presented in Section \ref{results}. In Section \ref{discussion}, we discuss the implications of our results, hypothesize on causes of the *A. raveneliana* decline, and suggest future research. Data and code are available at the study website (TODO: ADD LINK). 

# Materials and methods

## Study Specimens

Eighty-two *A. raveneliana* specimens were drawn from a single location in the Tuckasegee river. Ten were randomly selected and killed for baseline comparisons, and the remaining 72 were randomly assigned to 6 experimental sites (3 on each river) \@ref(fig:studyMap). The same number of *L. faciola* specimens were taken from the NC Wildlife Resources' Marian hatchery and were placed in a separate cage alongside the *A. raveneliana* as control specimens. The specimens were deposited at each site in April 2013. Due to high mortality at some sites, the *A. raveneliana* were pulled six months later. Some of the *L. faciola* were pulled at the same time, while the remaining were extracted the following Spring. 

```{r studyMap, echo = FALSE, fig.cap= "TODO: create a map of experimental sites."}
knitr::include_graphics(here::here("manuscript", "figures", "study_map.pdf"))
```

The mussels were held in plastic mesh cages anchored in the bottom substrate. All animals were individually identified with randomly selected numeric identification tags glued to the shells. Shell metrics and weight were obtained before and after deployment in the cages. After the experiment, a small (approximately 0.5 inch) section of the right valve from each animal was removed and embedded in plastic for laser ablation analyses.

## LA-ICP-MS

As youngest growth was of principal interest, we targeted growth parallel transects from the outer margins of valves, away from the umbo. For thicker valves we also analyzed additional offset scans to evaluate reproducibility and include older growth. In most cases, the entire preserved growth history was recorded by making multiple offset traverses (Figure \@ref(fig:alignment)). Along each transect, readings were taken every 2.88$\mu$m. The laser trajectories were drawn perpendicular to annular growth rings in a "step-scanning" approach [@shoults2014resolution], moving from the inner to the outer shell, beginning and ending in epoxy. The scans varied in terms of residing solely with the outer prismatic layer and periostracum, or being compound scans initiating in the nacreous layer. 

Each transect resulted in a chemical time-series for each analyte (concentration vs. distance; a matrix of about 7000 concentration measurements per mm). For data reduction (?) we assumed a stoichiometric CaCO3 composition of 39.547 wt\% Ca in nacreous tissue and 40.0432 wt\% Ca in other tissue and did not adjust for mineralogy (aragonite vs. calcite). We used USGS MACS-3 (synthetic aragonite) as the primary calibration standard, and analyzed NIST 614 (trace elements at ~1 ppm) and NIST 612 (trace elements at ~35 ppm). For drift compensation, each hour of data collection on unknowns was bracketed by analyses of each standard (in duplicate). We surveyed twenty industrial metals and bio-metals: Mg, Ca, Cr, Mn, Fe, Co, Ni, Cu, Zn, As, Se, Kr (for Se interference correction if detected), Sr, Mo, Cd, Sn, Ba, Hg, Pb and U. Analyses were conducted at the University of Texas at Austin using an ESI New Wave Research NWR193 excimer laser coupled to an Agilent 7500ce ICP-MS. 

## Anatomical Alignment

Review and annotation of high resolution cross section images was performed using Adobe Illustrator (Adobe Inc, San Jose CA). Annotations of the initial shell/epoxy boundary, annuli, nacre/prismatic layer boundary and terminal shell/epoxy boundary were made within Adobe Illustrator. Annuli were identified as any linear growth feature which could be traced from its origin in the umbo to its emergence through the prismatic and periostracum layers. Non-continuous linear features along the transect were not annotated and assumed to be “false annuli” caused by environmental disturbances [@veinott1996identification]. 

To align anatomical features with chemistry data, several methods were used to identify valve edges within the chemistry data based on algorithmic detection in mean and variance [@killick2014changepoint; @killick2016changepoint] or the minimum (or maximum) [@borchers2018pracma] of Ca43_CPS, Pb208_CPS, or the ratio of the two. The difference of the distance between these detected edges and the distance between the valves edges measured from high resolution scans was then computed. For each valve, the method that minimized this distance was then used to register the anatomy in the chemical data. Figure \@ref(fig:alignment) shows the process of anatomical alignment.

```{r alignment, echo = FALSE, fig.cap= "TODO"}
knitr::include_graphics(here::here("manuscript", "figures", "figure_alignment.pdf"))
```

## Chemical data pre-processing

Valves were grouped into seven batches for LA-ICP-MS processing. During each batch process, the lower limit of detection (LOD) was assessed multiple times using the standards described in Section \ref{la-icp-ms}. Concentration values were censored based on the mean LOD within a batch for each element. For analysis, parts-per-million (ppm) values were transformed to $\text{mol/Ca mmol}$ where the $Ca mmol$ was based on Ca\% of ~39.54\% for nacral observations and 40.078\% for all other layers.

## Study questions

We hypothesized that the distributions of one or more elements accrued in the nacre during the experiment would differ between experimental sites. To examine differential accrual, we posed a series of questions described in Table \ref{hypothesis-table}. The questions are designed to probe the primary research hypothesis by systematically examining both the sources and validity of any observed differences. 

Question (A) asks if there is difference any of the seven sites (baseline plus 3 experiment sites per river). Though notable, a difference between the baseline site and experimental sites does not improve our understanding of potential differences between the rivers, hence we then asked (B) whether experimental sites, excluding baseline, were different from each other and whether (C) the baseline site differed from the aggregated experimental sites. Questions (A-C) imply contrasts purposefully faithful to the completely randomized experimental design, thus permiting randomization-based inference [@splawa-neyman1990on-the-application; @coppock2019ri2] by permuting assignments to sites. 

If randomization is an unbiased instrument and we adqueately identified and linked elemental data to valve features, we should expect only random differences between nacral elemental distributions accrued *prior* to the experiment. Thus for question (D), we excluded observations from the most recent annuli, using pre-experiment observations as a negative control outcome [@arnold2016negative]. Lastly, in the case at least one experimental site is different in an element, we then explored trends within each river (E) and assocations with mordibity and mortality in the *A. raveneliana*. In the remaining subsections, the statistical methods are described in more detail.

\begin{table}
\label{hypothesis-table}
\begin{tabular}{llcccc}
Label & Question & Annuli included & \multicolumn{3}{c}{Methods} \\
 &   & & ri$_{gam}$ & ri$_{mom}$ & explore \\
\hline \\
A & Is at least one site (including baseline) different?  & All & yes & yes & \\
B & Is at least one site (excluding baseline) different?  & All & yes & yes & \\
C & Is the baseline site different from experiment sites? & All & yes & yes & \\
D & Is at least one site (including baseline) different?  & Excluding most recent & yes & yes & \\
E(a|b) & Is there trend within (Little Tennessee/Tuckasegee)? & All &  & & yes \\
\hline
\end{tabular}
\end{table}

## Statistical Analyses

We graphically summarized morbidity and mortality during the experiment by plotting trends in death and relative weight change at each site. 

### Outliers and sensitivity analyses

### Randomization-based inference

For questions A-D, null hypotheses of no difference between elemental distributions were tested using randomization inference in two ways, each of which make use of available data differently. Our experiment can be viewed as a cluster randomized trial, where the specimens are indepedent replicates and laser transects are clustered within a specimen. Inference was drawn using $m = 700$ simulations from the randomization distribution, using the `ri2` R package [@coppock2019ri2; @rcore2019]. The method can be summarized as follows: (1) compute the observed test statistic ($t_{obs}$) using the observed assignments, (2) permute the assignments by assigning specimens to different sites and re-compute the test statistic, (3) repeat (2) $m$ times. The two-sided p-value is the proportion of times $t_{obs}$ is greater than or equal to the test statistic distribution.

Though randomization inference preserves the nominal type I error level regardless of the chosen test statistic under the null hypothesis, power under various alternative hypotheses varies depending on this choice. We chose two statistics that used the available data in distinct ways with the goal to examine a range of possible diversions between nacral distributions be it in the shape of the trajectories or in distributional mean, variance, or extremes.

DESCRIBE stats:

* A test statistic based on the deviance between two nested GAM models preserved the temporal ordering of the observations.
* A statistic based on five summary statistics of each specimen's distributions. 

These statistics, the model specifications, and methodology are further described in the Supplementary Material.



#### L-moments

For descriptive purposes, we computed the proportion censored, first three probability of censor weighted L-moments ratios[@lmomco2018], and the maximum of observations within each anatomical area of interest for each element. These summary statistics were then plotted to examine trends in the elemental distributions.

# Results

## Morbidity and Mortality

Morbidity and mortality in *A. raveneliana* varied significantly across sites and between species (Figure \@ref(fig:mortality)). In both rivers, outcomes worsened downstream for *A. raveneliana*. Mortality increased monotonically down the river gradient, and the median relative weight change also tended to decrease. In the Little Tennessee, all but three specimens lost weight or died, whereas no individuals died and few lost weight at Tuckaseegee 1, the closest site to the baseline site. In constrast to *A. raveneliana*, none of the hatchery-raised and younger *L. fasciola* died and all gained weight.

```{r mortality, echo = FALSE, fig.cap= "TODO"}
knitr::include_graphics(here::here("manuscript", "figures", "mussel_outcomes_v000.pdf"))
```

## Trace element accrual

Figure \@ref(fig:accrualPvals) displays p-values for each hypothesis test described in Table XXXX. In *A. raveneliana*, Manganese shows the most evidence for the differences between sites among the elements. The baseline specimens do not appear different from the experimental specimens in terms of their Manganese trajectories (p = XXXX). Figure \@ref(fig:accrualPvals)B suggests that at least one experimental site is different from the others in terms of its Mn trajectories.

TODO: 

* discuss Pb
* discuss negative control outcome
* discuss trends (maybe?) seen in distributional summaries

In *L. fasciola*, Strontium and Barium show evidence of differences between sites. However, small p-value in Strontium appears to be driven by a difference between the baseline specimens and the specimens in experimental sites. This is in constrast to Barium, where the difference appears to be driven by difference among the experimental sites.

```{r accrualPvals, echo = FALSE, fig.cap= "TODO"}
knitr::include_graphics(here::here("figures", "ri_gam_ncr_pvals_v006.pdf"))
```

# Discussion

Based on current understanding of bivalve shell growth [@checa2000a-new-model], the laser transects that crossed the nacre should have captured the accretion that occurred during the experiment. By careful alignment of anatomic imaging and chemical data obtained by LA-ICP-MS, we were able to extract and analyze elemental concentrations from the nacre, including the locations of annuli within the nacre. We hypothesized that some elements would differentially accrue within the nacre and could be detected from LA-ICP-MS data. @jackson2003laser, for example, experimentally fed contaminated fish to banded water snakes and showed LA-ICP-MS used on tail tissue was able to differentiate between treatments.

Our analysis shows differences between the experimental sites in Mn concentrations accumulated during the study period. Manganese is an essential trace element with important roles in numerous enzymatic pathways that could have harmful effects on freshwater bivalves. We suggest two, possibly overlapping explanations for the Mn result. 

First, the differences in Manganese accumulation could have resulted from differences in survival times at the sites. @shoults2014resolution shows that Manganese concentrations naturally vary cyclically within the prismatic layer. The reasons for the seasonal variation in shell Mn are unknown. If Manganese also naturally cycles within the nacre, then shorter survival times would have resulted in different accrual patterns. Those that survived longer would have kept accumulating Mn for later in the season.

A second explanation for the Mn pattern is that the trajectories reflect differences in the water and/or substrate chemistries at the sites unrelated to seasonal variation. Interestingly, surviving valves from the Little Tennessee had microscopic crystal-like substance on their gill tissues, which upon mass spectrometry analysis consisted largely of Mn.

Other notes on Mn:

* correlation with growth bands: [@siegele2001manganese]
* toxic to snook: [@tuzuki2017effects]
* Mn bioaccumulation to mussels exposed to Mn: [@flores-montalvao2019how-leachates]
* "Mn negatively correlated with mussel survival and biomass": [@archambault2016assessing]
* "Valve movement responses to Mn were directly proportional to the activity of the free metal ion (Mn2 + ), which is consistent with the FIAM." [@markich2000valve]
* Mn in gills: @silverman1986interaction
* "The main results highlighted that only Mn caused a clear drop in glycogen levels in gills": @oliveira2019single
* Mn in archival monitoring: [@jeffree1995shell]
* @rzymski2014bioaccumulation
* higher densities of benthic invertebrates in Mn uncoated substrate: @wilson2019manganese
* pollution: @hameed1990freshwater, @wren1991the-effect

Notes on Mg:

* @kleinhenz2019acute

* TODO: look at soil and pollution maps. Do we see evidence of differential Mn at the sites?

These explanation point out limitations of our study design and measurements. The important scientific questions are related to the factors causing the *A. raveneliana*'s decline. The completely randomized design of our study permits internally valid causal inference in reference to the experimental assignments. Since we controlled elemental exposures only by proxy, nor do we have rich data of the water and substrate chemistries we cannot be robustly distinguish between the two explanations. Also, we do not know how long mussels survived during the experiment. Notching or otherwise marking the start of the experiment on the shells could have helped by measuring the amount valve deposited during the experiment [@haag2008testing].


The completely randomized design allows for robust, internally valid inference; however, further analyses with these data are needed to: 1) Examine other observed differences in elemental concentrations; 2) Rule out possible confounding such as seasonal and batch effects; and 3) Explore sources of these elements within the watersheds.

# References