---
title: Trace elements accrued in endangered freshwater bivalves in two North Carolina streams
author:
- Bradley C. Saul^[NoviSci, Inc, Durham, NC]
- Nathan R. Miller^[Department of Geological Sciences, University of Texas, Austin, TX, 78712-1692]
- Heather Gunn^[Department of Geological Sciences, University of Texas, Austin, TX, 78712-1692]
- Michael Walter^[Department of Marine, Earth and Atmospheric Sciences, North Carolina State University, Raleigh, NC, 27695-8208]
- Scott Salger^[School of Sciences, Barton College, Wilson,  NC, 27893]
- Mike Gangloff^[Department of Biology, Appalachian State  University, Boone, NC, 28608]
- Jay F. Levine^[Department of Marine, Earth and Atmospheric Sciences, North Carolina State University, Raleigh, NC, 27695-8208]
bibliography: elktoe.bib
toc: false
output: 
  bookdown::pdf_document2:
    number_sections: true
  bookdown::word_document2: default
header-includes:   
- \usepackage{amsmath, setspace}
- \usepackage{multirow}
- \graphicspath{ {figures/} } 
abstract: >
  \noindent Precipitous declines in the federally endangered Appalachian Elktoe (*Alasmidonta raveneliana*) have been observed since 2004 in the Little Tennessee River, but populations appeared stable in the Tuckasegee River in the same watershed. To assess river-specific factors associated with health and physiological outcomes, we conducted a completely randomized experiment, assigning specimens to sites within each river. Here, we examine differences between sites in trace element concentrations accumulated in specimens’ valves during the experiment.
keywords: TODO
---

\doublespacing

# Introduction 

Freshwater mussels (Bivalvia: Unionidae, Magaritifera) are predominately suspension feeding bivalves that inhabit streams, rivers and lakes.Unionids provide a range of ecosystem services [@vaughn2017ecosystem] They serve as food for raccoons, muskrats and other species that inhabit riparian habitat, and when abundant help stabilize streambeds. While feeding unionids filter the water column removing algae, bacteria and detritus and process and sequester some contaminants [@gagne2007changes, @vaughn2017ecosystem]. Unionid shells support a microecosytem of bacteria, protozoa and fungi that play a prominent role in stream ecology. Their shells are similar to tree rings, archiving some aspects of their environmental history, and have been used as bioindicators of pollution [@markich2002freshwater; @brown2005freshwater; @shoults2014resolution; @wilson2018freshwater]

North American freshwater mussel species are imperiled throughout their home range.  The majority, are federally or state listed as threatened, endangered or of special concern [@williams1993conservation]. Environmentally degrading riparian land-use practices have been attributed to their decline. Exposure to certain heavy metals from municipal effluents and other sources have been shown to increase morbidity and mortality in freshwater bivalves [@naimo1995a-review]. @hartmut2007declining attributed declining populations of the freshwater pearl mussel (*Margaritifera margaritifera*) to heavy metals and pesticides. Exposure to heavy metals have also been associated with genotoxic effects [@black1996dna, @khan2019bioaccumulation] enzymatic function, and behavioral changes [@naimo1995a-review].

The Appalachian Elktoe (*Alasmidonta raveneliana*) (Bivalvia: Unionidae) is native to the streams of eastern Tennessee and western North Carolina. The Elktoe is listed as federally endangered and remaining populations are found in the Nolichucky River in eastern Tennessee and the Upper Tennessee River system in North Carolina. Although once abundant in the Little Tennessee River, NC, precipitous declines were noted in  *A. raveneliana*  populations after major hurricane-related flooding associated with Hurricane Frances in 2004. However, within the same watershed’s  *A. raveneliana*  Tuckasegee River populations appeared stable. @miller2013concentrations evaluated sediment concentrations of Cu, Cr, Ni, and Zn in the Little Tennessee and Tuckasegee. When considered with the chemical concentrations of *A. raveneliana* shells measured by @jarvis2011water, they found these elements (Cu in particular) unlikely to be the primary culprit in the *A. raveneliana* decline. Reasons for the decline remain unclear [@jarvis2011water; @miller2013concentrations; miller2016potential; @pandolfi2016effects]. 

To determine how river-specific factors may be associated with health and physiological outcomes, we conducted a completely randomized experiment by assigning eighty-two  *A. raveneliana*  along with an equal number of hatchery-raised Wavy-Rayed Lampmussel (*Lampsilis fasciola*) to sites within each river [@salger2020]. In this paper, we ask whether exposure to the experimental sites modified the accretion of twenty trace elements in specimens’ valves. We demonstrate how chemical trajectories obtained by laser ablation inductively coupled plasma mass spectrometry (LA-ICP-MS) transects [@pozebon2014review] can be aligned with anatomical features of the valves, thus making comparisons of specific valve layers and time periods possible. 

# Materials and methods

## Study Specimens

Eighty-two *A. raveneliana* specimens were drawn from a single location in the Tuckasegee river [@salger2020]. Ten were randomly selected and sacrificed for baseline comparisons, and the remaining 72 were randomly assigned to 6 experimental sites, three each in the Tuckasegee and Little Tennessee Rivers (Figure \@ref(fig:studyMap)). The same number of young adult *L. faciola* specimens were obtained from the NC Wildlife Resources' Marian hatchery and randomly assigned to a separate set of cages at the six sites. The plastic mesh cages were anchored in the bottom substrate. All animals were individually identified with randomly selected numeric identification tags glued to the shells. Shell metrics and weight were obtained before and after deployment in the cages. After the experiment, a small (approximately 0.5 inch) section of the right valve from each animal was removed and embedded in plastic for laser ablation analyses.

```{r studyMap, echo = FALSE, fig.cap= "TODO: create a map of experimental sites."}
knitr::include_graphics(here::here("manuscript", "figures", "study_map.pdf"))
```

## LA-ICP-MS

Since the study was initiated to identify factors potentially contributing to the decline of the system's Elktoe populations, our analysis focused on the sections of shell that reflected areas of shell deposition during the period of the sentinel experiment. Parallel transects from the outer margins of valves, away from the umbo were targeted for analysis. For thicker valves we also analyzed additional offset scans to evaluate reproducibility and include older growth in the prismatic layers. In most cases, the entire preserved growth history was recorded by making multiple offset traverses (Figure \@ref(fig:alignment)). Along each transect, readings were taken every 2.88$\mu$m. The laser trajectories were drawn perpendicular to annular growth rings in a "step-scanning" approach [@shoults2014resolution], moving from the inner to the outer shell, beginning and ending in epoxy. Scans varied depening on location, in terms of residing solely with the outer prismatic layer and periostracum, or being compound scans initiating in the nacreous layer. 

Each transect resulted in a chemical time-series for each analyte (concentration vs. distance; a matrix of about 7000 concentration measurements per mm). For data reduction^[Nathan/Heather - can you explain what this means?] we assumed a stoichiometric CaCO3 composition of 39.547 wt\% Ca in nacreous tissue and 40.0432 wt\% Ca in other tissue and did not adjust for mineralogy (aragonite vs. calcite). We used USGS MACS-3 (synthetic aragonite) as the primary calibration standard, and analyzed NIST 614 (trace elements at ~1 ppm) and NIST 612 (trace elements at ~35 ppm). For drift compensation, each hour of data collection on unknowns was bracketed by analyses of each standard (in duplicate). We surveyed the shells for twenty industrial metals and bio-metals: Mg, Ca, Cr, Mn, Fe, Co, Ni, Cu, Zn, As, Se, Kr (for Se interference correction if detected), Sr, Mo, Cd, Sn, Ba, Hg, Pb and U. Analyses were conducted at the University of Texas at Austin using an ESI New Wave Research NWR193 excimer laser coupled to an Agilent 7500ce ICP-MS. 

## Anatomical Alignment

Review of high resolution cross section images and annotation of the initial shell/epoxy boundary, annuli, nacre/prismatic layer boundary and terminal shell/epoxy boundary were made using Adobe Illustrator (Adobe Inc, San Jose CA). Annuli were identified as any linear growth feature which could be traced from its origin in the umbo to its emergence through the prismatic and periostracum layers. Non-continuous linear features along the transect were not annotated and assumed to be “false annuli” caused by environmental disturbances [@veinott1996identification]. 

Several methods were used to identify valve edges and align anatomical features with chemistry data based on algorithmic detection in mean and variance [@killick2014changepoint; @killick2016changepoint] or the minimum (or maximum) [@borchers2018pracma] of Ca43_CPS, Pb208_CPS, or the ratio of the two (Figure \@ref(fig:alignment)). The difference of the distance between these detected edges and the distance between the valves edges measured from high resolution scans was then computed. For each valve, the method that minimized this distance was then used to locate anatomical locations in the chemical data. 

```{r alignment, echo = FALSE, fig.cap= "TODO"}
knitr::include_graphics(here::here("manuscript", "figures", "figure_alignment.pdf"))
```

## Chemical data pre-processing

Valves were grouped into seven batches for LA-ICP-MS processing. During each batch process, the lower limit of detection (LOD) was assessed multiple times using the standards noted above. Concentration values were left-censored based on the mean LOD within a batch for each element. For analysis, parts-per-million (ppm) values were rescaled to $\text{mol/Ca mmol}$ where $\text{Ca mmol}$ was based on Ca\% of ~39.54\% for nacral observations and 40.078\% for all other layers.

## Study questions

Several statistical measures were used to probe distributions of elements in the nacre that accumulated both before and during the experiment to determine if any of the elements showed a differential accretion. As assignment to sites was completely randomized, we used the randomization distribution [@coppock2019ri2] to draw inference to changes in the elemental distributions. We hypothesized that the distributions of one or more elements accrued in the nacre during the experiment would differ between experimental sites. To examine differential accrual, we probed the primary research hypothesis by systematically examining both the sources and validity of any observed differences in shell composition in manner that was faithful to the completely randomized experimental design, thus permiting randomization-based inference [@splawa-neyman1990on-the-application; @coppock2019ri2] by permuting assignments to sites. 

Systematic differences in nacral elements accrued *prior* to the experiment could bias differences accrued during the experiment. Thus, to check that the sources of variation in the experiment were not due pre-experiment differences we first asked if there were differences in elemental concentrations *excluding* nacre accrued during the experiment. We examined this question both with (Q1) and without (Q2) baseline specimens which were not exposed to the experiment. Our primary focus was on whether nacre accrued during the experiment showed differences in elemental concentrations among and between experimental sites (Q3). 

## Statistical Analyses

The experiment was viewed as a cluster randomized trial, where the specimens were indepedent replicates and LA-ICP-MS transects were clustered within a specimen. Null hypotheses of no difference between elemental distributions were tested using randomization-based inference using $m = 1000$ simulations from the experiment's distribution. Each analysis was conducted in the following series of steps: (1) the observed test statistic ($t_{obs}$) was computed using the observed assignments, (2) assignments were permuting by assigning specimens to different sites and test statistics re-computed, (3) repeat (2) $m$ times. The two-sided p-value was the proportion of times $t_{obs}$ was greater than or equal to the test statistic's distribution. The `ri2` R package was used to carry out inference [@coppock2019ri2; @rcore2019].

Though randomization inference preserves the nominal type I error level under the null hypothesis regardless of the chosen test statistic, power under various alternative hypotheses may vary depending on this choice. Consquently, we chose two test statistics (Table \ref{test-statistics}) that use the available data in different ways and are able to test different ways in which the distributions of concentrations may have changed. 

The first test statistic was the deviance between two nested GAM models (labelled ri$_{gam}$), one which included terms for experimental sites and a nested model which did not. For the ri$_{gam}$ analyses, we used the one transect per specimen which had the most number observations from the most recent annuli. The specification of the GAM models preserved the ordering of the observations along the transect within the nacre. The GAM models did not, however, account for censoring for observations below a lower limit of detection.

The second test statistic (ri$_{wls}$) was an $F$-statistic comparing two nested weighted least squares regressions. Similar to ri$_{gam}$, the nested model did not include terms for the experimental sites. The dependent variables were each specimen's mean element concentration from observations from the youngest annuli for a given question. For example, for questions (Section \ref{study-questions}) involving the experimental period the youngest of all annuli was used; however, for questions regarding the pre-experiment time, the second youngest annuli was used. Weights in the regression were the sample variance taken over the same observations used to compute the mean.

All the models used in ri$_{gam}$ and ri$_{wls}$ included terms for a specimen's volume (height $\times$ width $\times$ length) measured at baseline and its LA-ICP-MS batch.

\begin{table}
\label{test-statistics}
\caption{Data and test statistics used for randomization inference}
\begin{tabular}{lp{5cm}p{7cm}}
Label & Transect Data Used & Statistic(s) \\
\hline \\
$ri_{\text{gam}}$ & transect with the most observations from the youngest annuli & deviance between nested GAM models \\
$ri_{\text{wls}}$  & all transects with any youngest annuli observations  & $F$-statistic between nested WLS models \\
\hline
\end{tabular}
\end{table}

### Sensitivity analyses

The point of transition from epoxy to shell was a potential source of error (Section \ref{anatomical-alignment}), thus we checked the stability of our results for possible misclassification of nacral observations as epoxy by repeating the analysis after trimming 6$\mu$m and 12$\mu$m from the nacral observations. We also repeated analyses after preprocessing transects by truncating values to 3 standard deviations from a moving average filter [@signal2014].

The methodology, test statistics, and the model specifications are further described in the Supplementary Material.

# Results

The experimental p-values corresponding to hypothesis tests for Q1-Q3 for each element are shown in Figure \@ref(fig:rires). Barium in *L. fasciola* is the only element with p-values suggesting strong significant differences based on both test statistics. However, since the differences appear for Q1 and Q2 (Figure \@ref(fig:rires)A-B), the result for Q3 (Figure \@ref(fig:rires)C) is likely due to pre-existing differences.

```{r rires, echo = FALSE, fig.cap = "Points are p-values (log$_{10}$ scale) testing the null hypothesis of no difference in each element accrued prior to the experiment between (A) sites including baselines excluding most recent annuli, (B) sites excluding baseline excluding most recent annuli, (C) sites excluding baseline including most recent annuli. Smaller p-values suggest a larger discrepancy between sites under a null hypothesis of no difference between sites. P-values $<0.1$ are labelled by element."}
knitr::include_graphics(here::here("manuscript", "figures", "ri_ncr_hypothesis_v001.pdf"))
```

Figure \@ref(fig:risummary) shows test statistic summaries for elements where both test statistic p-values are $<0.1$ for the main study question (Q3). In *A. raveneliana*, this includes elements Co, Cu, and Mn; in *L. fasciola*, this includes Ba. Figure \@ref(fig:risummary) reveals that the differences in *L. fasciola* Ba concentrations are due to low concentrations among specimens from the first Little Tennessee site. Differences in *A. raveneliana* Mn concentration are due to lower values in the first Tuckasegee site. 

```{r risummary, echo = FALSE, fig.cap = "Plots examine test statistics in cases where both $ri_{wls}$ and $ri_{gam}$ had p-values $<0.1$ in testing the hypothesis of differences among sites accrued during the experiment. (A) Point are predicted inverse-variance weighted mean concenrations in the youngest annuli. (B) Lines are GAM smoothed estimates of the nacral elemental trajectories by site, showing the first 200 $\\mu$m starting from the innder edge of the nacre."}
knitr::include_graphics(here::here("manuscript", "figures", "ri_ncr_summary_v001.pdf"))
```

# Discussion

Based on current understanding of bivalve shell growth [@checa2000a-new-model], the laser transects that crossed the nacre should have captured the accretion that occurred during the experiment. By careful alignment of anatomic imaging and chemical data obtained by LA-ICP-MS, we were able to extract and analyze elemental concentrations from the nacre, including the locations of annuli within the nacre. We hypothesized that some elements would differentially accrue within the nacre across experimental assignments and could be detected from LA-ICP-MS data. @jackson2003laser, for example, experimentally fed contaminated fish to banded water snakes and showed LA-ICP-MS used on tail tissue was able to differentiate between treatments.

Our analysis revealed no clear cut differences between sites that could be easily explained by the experiment itself. The lower concentrations of Barium in Little Tennesse Site 1 *L fasciola* specimen has no obvious reason, for example. All *L fasciola* specimens were hatchery raised in the same conditions. The shells for the experimental sites were randomly distributed in the LA-ICP-MS sequence. 

Several factors complicate interpretation of our results: censoring due to death, inaccurate identification experiment time within the valve, incomplete understanding of physiological processes especially temporal stationarity of elements. To explain these factors, we consider the differences between the experimental sites in Mn concentrations accumulated during the study period.

Manganese is an essential trace element with important roles in numerous enzymatic pathways that could have harmful effects on freshwater bivalves. Interestingly, surviving valves from the Little Tennessee had films on the gills that were apparent by scanning electron microscopy, which upon x-ray diffraction analysis consisted largely of Mn [@levine2020].

Manganese valve concentrations have been shown to correlate with seasonal changes [@siegele2001manganese; @shoults2014resolution]. The reasons for the seasonal variation are unknown, but in our study, different survival times could have resulted in different accrual patterns. Those that survived longer would have kept accumulating Mn for later in the season. We do not know when specimens in the experiment died. It could be that those specimens that survived to the end of the study (Tuckasegee 1 site had the lowest mortality) were accruing shell for longer while the trajectories for other specimens were censored at their time of death.

The problem of trajectories censored by death is related to a similar problem. We assumed that the youngest annuli in the nacre accrued during the experiment. We don't know for sure, and we have no way of testing this. In Little Tennessee specimens, for example, where high morbidity and mortality was observed in *A. raveneliana*, 
it could be that little to no nacre accrued during the experiment. This situation would invalidate our experimental comparisons. Notching or somehow marking the start of the experiment on the shells could have helped track more precisely which parts of the value were deposited during the experiment [@haag2008testing].

TODO:

* discuss last point on "incomplete understanding of physiological processes": are elements labile within nacre or primsatic layer or are they conserved?
* wrap up

# Funding

This work was supported by funds provided by the North Carolina Department of Transportation [NCDOT Project 2013-2012-37]; and the US Fish and Wildlife Service [F18AC00237].

# Acknowledgment

The authors thank [TODO]


# References