---
title: No clear differences in trace elements accrued in endangered freshwater bivalves randomized to two North Carolina streams
author:
- Bradley C. Saul^[NoviSci, Inc, Durham, NC]
- Nathan R. Miller^[Department of Geological Sciences, University of Texas, Austin, TX, 78712-1692]
- Heather Gunn^[Department of Geological Sciences, University of Texas, Austin, TX, 78712-1692]
- Michael Walter^[Department of Marine, Earth and Atmospheric Sciences, North Carolina State University, Raleigh, NC, 27695-8208]
- Scott Salger^[School of Sciences, Barton College, Wilson,  NC, 27893]
- Mike Gangloff^[Department of Biology, Appalachian State  University, Boone, NC, 28608]
- Jay F. Levine^[Department of Marine, Earth and Atmospheric Sciences, North Carolina State University, Raleigh, NC, 27695-8208]
bibliography: elktoe.bib
toc: false
output: 
  bookdown::pdf_document2:
    number_sections: true
  bookdown::word_document2: default
header-includes:   
- \usepackage{amsmath, setspace, caption, longtable}
- \usepackage{multirow}
- \graphicspath{ {figures/} } 
abstract: >
  \noindent Precipitous declines in the federally endangered Appalachian Elktoe (*Alasmidonta raveneliana*) have been observed since 2004 in the Little Tennessee River, but populations appeared stable in the Tuckasegee River in the same watershed. To assess river-specific factors associated with health and physiological outcomes, we conducted a completely randomized experiment, assigning specimens to three sites within each river. High morbidity and mortality was observed in certain sites. We examine associations between twenty trace elements in valves both with experimental sites and with health outcomes. 
keywords: TODO
---

\doublespacing

```{r, echo = FALSE, message=FALSE}
library(elktoeChemistry)
library(gt)
```

# Introduction 

Freshwater mussels (Bivalvia: Unionidae, Magaritifera) are predominately suspension feeding bivalves that inhabit streams, rivers and lakes. Unionids provide a range of ecosystem services [@vaughn2017ecosystem]. They serve as food for raccoons, muskrats and other species that inhabit riparian habitat, and when abundant help stabilize streambeds. While feeding unionids filter the water column removing algae, bacteria and detritus and process and sequester some contaminants [@gagne2007changes; @vaughn2017ecosystem]. Unionid shells support a microecosytem of bacteria, protozoa and fungi that play a prominent role in stream ecology. Their shells are similar to tree rings, archiving some aspects of their environmental history, and have been used as bioindicators of pollution [@markich2002freshwater; @brown2005freshwater; @shoults2014resolution; @wilson2018freshwater]

North American freshwater mussel species are imperiled throughout their home range. The majority are federally or state listed as threatened, endangered or of special concern [@williams1993conservation]. Environmentally degrading riparian land-use practices have been attributed to their decline. Exposure to certain heavy metals from municipal effluents and other sources have been shown to increase morbidity and mortality in freshwater bivalves [@naimo1995a-review]. @hartmut2007declining attributed declining populations of the freshwater pearl mussel (*Margaritifera margaritifera*) to heavy metals and pesticides. Exposure to heavy metals have also been associated with genotoxic effects [@black1996dna; @khan2019bioaccumulation] enzymatic function, and behavioral changes [@naimo1995a-review].

The Appalachian Elktoe (*Alasmidonta raveneliana*) (Bivalvia: Unionidae) is native to the streams of eastern Tennessee and western North Carolina. The Elktoe is listed as federally endangered and remaining populations are found in the Nolichucky River in eastern Tennessee and the Upper Tennessee River system in North Carolina. Although once abundant in the Little Tennessee River, precipitous declines were noted in  *A. raveneliana*  populations after major hurricane-related flooding associated with Hurricane Frances in 2004. However, within the same watershed’s *A. raveneliana* Tuckasegee River populations appeared stable. @miller2013concentrations evaluated sediment concentrations of Cu, Cr, Ni, and Zn in the Little Tennessee and Tuckasegee. When considered with the chemical concentrations of *A. raveneliana* shells measured by @jarvis2011water, they found these elements (Cu in particular) unlikely to be the primary culprit in the *A. raveneliana* decline. Reasons for the decline remain unclear [@jarvis2011water; @miller2013concentrations; @miller2016potential; @pandolfi2016effects]. 

To determine how river-specific factors may be associated with health and physiological outcomes, we conducted a completely randomized experiment by assigning eighty-two *A. raveneliana* along with an equal number of hatchery-raised Wavy-Rayed Lampmussel (*Lampsilis fasciola*) to sites within each river. In this paper, we explore associations between twenty trace elements in specimens’ valves and experimental sites as well as associations with health outcomes. We demonstrate how chemical trajectories obtained by laser ablation inductively coupled plasma mass spectrometry (LA-ICP-MS) transects of valve cross-sections [@pozebon2014review] can be aligned with anatomical features of the valves in order to compare specific valve layers and time periods. 

# Materials and methods

## Study Specimens

Eighty-two *A. raveneliana* specimens were collected from a single location in the Tuckasegee river. Ten were randomly selected and sacrificed for baseline comparisons, and the remaining 72 were randomly assigned and relocated in cages to six experimental sites, three each in the Tuckasegee and Little Tennessee Rivers (see appendix for site locations). The same number of young adult *L. faciola* specimens, obtained from the NC Wildlife Resources' Marian hatchery, were randomly assigned to a separate cages at the six sites. The plastic mesh cages were anchored in the bottom substrate. All animals were individually identified with randomly selected numeric identification tags glued to the shells. Weight and other metrics were obtained before and after deployment in the cages. After the experiment, a small (approximately 0.5 inch) section of the right valve from each animal was removed and embedded in epoxy for laser ablation analyses.

Four specimens were removed from analysis due to anomalies in their valve anatomy or in the laser transects.

## LA-ICP-MS

Since the study was initiated to identify factors potentially contributing to the decline of the Elktoe populations, our analysis focused on the sections of shell that biomineralized during the period of the sentinel experiment. Parallel transects from the outer margins of valves, away from the umbo were targeted for analysis. For thicker valves we also analyzed additional offset scans to evaluate reproducibility and include older growth in the prismatic layers. In most cases, the entire preserved growth history was recorded by making multiple offset traverses (Figure \@ref(fig:alignment)). We used a 25$\mu$m round aperture moving at 5$\mu$m/sec. The quadrupole sampling period (0.5762 s) enabled a moving compositional average for the 20 surveyed analytes every 2.88 $\mu$m (8.7 readings per 25$\mu$m footprint). For valve cross-sectional thicknesses on the order of 1 mm, transects included roughly 350 measurements. The laser trajectories were drawn perpendicular to annular growth rings in a "step-scanning" approach [@shoults2014resolution], moving from the inner to the outer shell, beginning and ending in epoxy. Scans varied depending on location, in terms of residing solely with the outer prismatic layer and periostracum, or being compound scans initiating in the nacreous layer. 

Each transect resulted in a chemical time-series for each analyte (concentration vs. distance; a matrix of about 7000 concentration measurements per mm). For data reduction^[Nathan/Heather - can you explain what this means?] we assumed a stoichiometric CaCO3 composition of 39.547 wt\% Ca in nacreous tissue and 40.0432 wt\% Ca in other tissue and did not adjust for mineralogy (aragonite vs. calcite){>>Include EPMA work done to evaluate Ca concentrations of the nacreous and prismatic layers.  The nacreous layer compositions would need a slight correction for the difference in Ca concentration (used as the internal standard).<<}. We used USGS MACS-3 (synthetic aragonite) as the primary calibration standard, and analyzed NIST 614 (trace elements at ~1 ppm) and NIST 612 (trace elements at ~35 ppm). For drift compensation, each hour of data collection on unknowns was bracketed by analyses of each standard (in duplicate). We surveyed the shells for twenty industrial metals and bio-metals: Mg, Ca, Cr, Mn, Fe, Co, Ni, Cu, Zn, As, Se, Kr (for Se interference correction if detected), Sr, Mo, Cd, Sn, Ba, Hg, Pb and U. Analyses were conducted at the University of Texas at Austin using an ESI New Wave Research NWR193 excimer laser coupled to an Agilent 7500ce ICP-MS. 

## Anatomical Alignment

Review of high resolution cross section images and annotation of the initial shell/epoxy boundary, annuli, nacre/prismatic layer boundary and terminal shell/epoxy boundary were made using Adobe Illustrator (Adobe Inc, San Jose CA). Annuli were identified as any linear growth feature which could be traced from its origin in the umbo to its emergence through the prismatic and periostracum layers. Non-continuous linear features along the transect were not annotated and assumed to be "false annuli" caused by environmental disturbances [@veinott1996identification]. 

To align anatomical features with chemistry data, several algorithms were used to first identify a reference point such as the epoxy/nacre boundary in the chemistry data. These methods included detecting changes in mean, variance [@killick2014changepoint; @killick2016changepoint] or the extrema [@borchers2018pracma] of Ca43_CPS, Pb208_CPS, or the ratio of the two (Figure \@ref(fig:alignment)). The difference of the distance between the detected reference point and the distance between the valves edges measured from high resolution scans was then computed. For each valve, the method that minimized this distance was then used to locate anatomical locations in the chemical data. The project website includes the code and complete details of the alignment steps.

```{r alignment, echo = FALSE, fig.cap= "TODO"}
knitr::include_graphics(here::here("manuscript", "figures", "figure_alignment.pdf"))
```

## Chemical data pre-processing and de-noising

Valves were grouped into seven batches for LA-ICP-MS processing. During each batch process, the lower limit of detection (LOD) was assessed multiple times using the standards noted above. Concentration values were left-censored based on the mean LOD within a batch for each element. For analysis, parts-per-million (ppm) values were rescaled to $\text{mol/Ca mmol}$ where $\text{Ca mmol}$ was based on Ca\% of ~39.54\% for nacral observations and 40.078\% for all other layers. Chemical signals were denoised by a rolling median absolute deviation filter with a bandwidth of ten observations.

## Study Questions

We hypothesized that the distributions of one or more elements accrued in valve layers would differ between experimental sites. We examined differences between sites within the following valve layers: the youngest annuli in the nacre, older nacre, prismatic layer, and the periostracum. The epoxy layers were also included in the analysis as a negative control -- we should should see no difference between sites in epoxy. 

In the case that differences between sites were observed, we then looked at associations between health outcomes and elemental concentrations. These examinations are exploratory as associations between health and elements that accrued post-randomization cannot be interpreted causally without strong, untestable assumptions.

## Statistical Analyses

Specimens' weight and size measurements at the start of the experiment were summarized by site with means and standard deviations. Morbidity in terms of relative weight change and mortality were graphically displayed to highlight any trends among the experimental sites.

The distribution of elements within each layer are summarized by its central tendency. Due to the high degree of value below detection levels in some elements, we estimated the first L-moment based on probability weighted moments for left censored data [@lmomco2018]. In the case that all values for a specimen were below detection within a layer, the detection limit was used as the layer's summary statistic. L-moments are computed from order statistics and are less sensitive to outliers compared to standard moment estimates such as the arithmetic mean.

Median and interquartile ranges of the moment estimates were tabulated by species, layer, and site. Within each species/layer combination, the null hypothesis of no difference between any site was tested by the nonparametric Kruskal-Wallis method [@hollander2013nonparametric]. To account for possible batch effects in LA-ICP-MS processing, tests used the residuals from a no-intercept linear model including a term for the LA-ICP-MS drawer. The tests were done both including and excluding the baseline site. No control for type 1 error was done. P-values are presented as descriptive measures, hence no statistical significance threshold was set. Analyses were done using R version `r paste(R.Version()$major, R.Version()$minor, sep = ".")` [@rcore2020]. 

# Results

Table XX shows characteristics of the specimens at the start of the experiment. Figure \@ref(fig:outcomes) displays morbidity and relative weight changes in the experimental sites. In *A. raveneliana*, morbidity and mortality increased in a downstream direction in both rivers, though outcomes were worse in the Little Tennessee. None of the younger *L. fasciola* died and all gained weighted during the experiment.

```{r table1, results = 'asis', echo=FALSE}
x <- readLines(here::here("manuscript", "figures", "table1.tex"))
x <- x[!grepl("\\\\small ", x)]
x <- c(x[1], "\\label{tab:table1}", x[2:length(x)])
cat(x)
```

```{r outcomes, echo = FALSE, fig.cap= "TODO"}
knitr::include_graphics(here::here("manuscript", "figures", "mussel_outcomes_v000.pdf"))
```

Summaries of element distributions by layer and site are shown in Table XX. P-values from the Kruskal-Wallis tests are displayed in Figure \@ref(fig:pvalues) for all elements. The x-axis is on the log 10 scale, hence smaller p-values are to the right. Tests were based on residuals from a simple linear model adjusting for LA-ICP-MS drawer to account for possible batch effects. Results are shown both including baseline sites and using only the experimental sites.

```{r element_table, results = 'asis', echo=FALSE}
x <- readLines(here::here("manuscript", "figures", "element_summary.tex"))
x <- x[!grepl("\\\\small ", x)]
x <- c(x[1], "\\label{tab:summary}", x[2:length(x)])
cat(x)
```


```{r pvalues, echo = FALSE, fig.cap= "Kruskal-Wallis test p-values from a testing the null hypothesis of no difference between sites using the residuals from a simple linear model adjusting for LA-ICP-MS drawer."}
knitr::include_graphics(here::here("manuscript", "figures", "pvalues.pdf"))
```

Including the baseline site tended to result in smaller p-values. In *A. raveneliana* all p-values for all elements were greater than 0.001, and the smallest p-value for this species for the experiment sites was 0.1 (Barium). Also in *A. raveneliana*, Smaller p-values in comparisons including the baseline site tended to occur in the prismatic layer.

The *L. fasciola* with baseline sites have p-values less than 0.0001 for 5 element/layer combinations (Ba and Sr - periostracum and prismatic layers, Mn - prismatic layer). Among the *L. fasciola* from experimental sites only prismatic layer Ba and Mn show strong evidence against the null hypothesis.

# Discussion

Based on current understanding of bivalve shell growth [@checa2000a-new-model], the laser transects that crossed the nacre should have captured the accretion that occurred during the experiment. By aligning of anatomic images and chemical data obtained by LA-ICP-MS, we were able to extract and analyze elemental concentrations from the nacre, including the locations of annuli within the nacre. We sought to determine whether elemental concentrations obtained from LA-ICP-MS differed across experimental assignments in each bivalve layer, especially in the nacral annuli we classified as the youngest. @jackson2003laser, for example, experimentally fed contaminated fish to banded water snakes and showed LA-ICP-MS used on tail tissue was able to differentiate between treatments.

Our study expands on existing research on *A. raveneliana* [@miller2013concentrations] by expanding the number of trace elements examined. To our knowledge, it is also the only completely randomized experiment. Our analysis revealed no clear cut differences between sites that could be easily explained by the experiment itself. The lower concentrations of Barium in Little Tennessee Site 1 *L. fasciola* specimena has no obvious explanation, for example. All *L. fasciola* specimens were hatchery raised in the same conditions. The shells for the experimental sites were randomly distributed in the LA-ICP-MS sequence, suggesting the result is not an artifact of LA-ICP-MS processing. Moreover, *A. raveneliana* specimens do show a similar pattern in the nacre. Since *A. raveneliana* were much older than *L. fasciola*, comparing prismatic layer difference across species may not be relevant, as the laser transect in general crossed older prismatic layer sections in *A. raveneliana*. Moreover, most of the *L. fasciola* were held at the experimental sites six months after the *A. raveneliana*, allowing this species to experience a different season and longer times.

We were particularly interested in Mn, as it is an essential trace element with important roles in numerous enzymatic pathways that could have harmful effects on freshwater bivalves. Interestingly, surviving valves from the Little Tennessee had films on the gills that were apparent by scanning electron microscopy, which upon x-ray diffraction analysis consisted largely of Mn.

Manganese valve concentrations have been shown to correlate with seasonal changes [@siegele2001manganese; @shoults2014resolution]. The reasons for the seasonal variation are unknown, but in our study, different survival times could have resulted in different accrual patterns. Those that survived longer would have kept accumulating Mn and other elements for later in the season. We do not know when specimens in the experiment died. It could be that those specimens that survived to the end of the study (Tuckasegee 1 site had the lowest mortality) were accruing shell for longer while the trajectories for other specimens were censored at their time of death.

We assumed that the youngest annuli in the nacre accrued during the experiment, but the data provide no way of testing this assumption. In Little Tennessee specimens, where high morbidity and mortality was observed in *A. raveneliana* it could be that little to no nacre accrued during the experiment. Marking the start of the experiment shells in some way may have provided appropriate data, but @haag2008testing points out that notching shells does harm individuals. Though we carefully aligned valve layers with chemistry data, in some cases layers were most certainly misclassified particularly at layer boundaries. Future studies should consider ways of safely marking the start of the experiment on the valves in order to measure and classify the valve that accrued during the experiment.

The causes for *A. raveneliana* declines remain a mystery, as our results do not show any clear association between contaminants and experiment sites or health outcomes. The articles section of the project website [bsaul.github.io/elktoeChemistry/](https:://bsaul.github.io/elktoeChemistry/) contains additional analyses and project details. Data and code for this project are available from the project's github repository [github.com/bsaul/elktoeChemistry](https://github.com/bsaul/elktoeChemistry).

# Funding

This work was supported by funds provided by the North Carolina Department of Transportation [NCDOT Project 2013-2012-37]; and the US Fish and Wildlife Service [F18AC00237].

# Acknowledgment

The authors thank [TODO]


# References