---
title: Trace elements accrued in endangered freshwater bivalves in two North Carolina streams
author:
- Bradley C. Saul^[NoviSci, Inc, Durham, NC]
- Nathan R. Miller^[Department of Geological Sciences, University of Texas, Austin, TX, 78712-1692]
- Heather Gunn^[Department of Geological Sciences, University of Texas, Austin, TX, 78712-1692]
- Michael Walter^[Department of Marine, Earth and Atmospheric Sciences, North Carolina State University, Raleigh, NC, 27695-8208]
- Scott Salger^[School of Sciences, Barton College, Wilson,  NC, 27893]
- Mike Gangloff^[Department of Biology, Appalachian State  University, Boone, NC, 28608]
- Jay F. Levine^[Department of Marine, Earth and Atmospheric Sciences, North Carolina State University, Raleigh, NC, 27695-8208]
bibliography: elktoe.bib
toc: false
output: 
  bookdown::pdf_document2:
    number_sections: true
  bookdown::word_document2: default
header-includes:   
- \usepackage{amsmath, setspace}
- \usepackage{multirow}
- \graphicspath{ {figures/} } 
abstract: >
  \noindent Precipitous declines in the federally endangered Appalachian Elktoe (*Alasmidonta raveneliana*) have been observed since 2004 in the Little Tennessee River, but populations appeared stable in the Tuckasegee River in the same watershed. To assess river-specific factors associated with health and physiological outcomes, we conducted a completely randomized experiment, assigning specimens to sites within each river. Here, we examine differences between sites in trace element concentrations accumulated in specimens’ valves during the experiment.
keywords: TODO
---

# TODO List (as of `r Sys.Date()`) {-}

* [ ] clean up map; is it BW ready? Better color of labels. Remove lat/lon labels. Fewer lat/lon breaks. caption
* [ ] update figure (Figure 2) describing data processing. crop valve. remove "step 3". Show GAM fitting for single element.
* [ ] develop online supplement (bsaul.github.io/elktoeChemistry)
  * [x] should contain: data preparation and analysis code
  * [ ] detailed description of LA-ICP-MS
  * [ ] any add'l analyses
* [ ] text edits/updates
  * develop results section and ensure that it aligns with discusson
  * discuss design of transects for future studies: fly-scanning along prismatic layer and/or (if possible) some sort of probabilistic sampling within each valve (even simply a systematic sample every X units from umbo to mantle.)
* [x] push code to github
* [ ] push data to github
* [ ] format manuscript for target journal

# NM comments {-}

*  What is current status of A. Rav?
*  Perhaps specify the research hypotheses investigated in this study.  
    1.	 Assess the compositional differences in shell composition as a function of specimen health/morbidity and intra-watershed location.
    2.	 Among these populations:
        a.	 compare inner (nacreous) to outer (prismatic layers) shell compositions during the experimental period.
        b.	 Compare inner to outer shell compositions preceding the experimental period, as sampled by replicate whole-valve transects
        away from the umbo, to assess whether innermost and/or outermost valve layers, biomineralized prior to the experiment, were
        affected by environmental exposure associated with relocation.
* Need to say how old they were; even if retroactive from the annuli counting.  How does the experimental period duration compare to their ages?  Where do we expect to see growth during the experiment? Since seasonality if a factor for shell growth, when did the experiment take place in this regard? How many specimens had traverses that spanned nacreous AND prismatic calcite deposition during the experiment? These are helpful stats for assessing the population from which inferences are made.
* On figure 2: possibly include predominant land use and location of any industrial sites on this map?  What are grey areas?  What is lake?  Could bedrock geology also be shown?
* On LOD: Vague; For laser traverses, there is little control over baseline values during the traverse.  Accordingly for travers, the Limits of detection are inferred from the analysis of standards analyzed ~hourly over the analytical session.  These are typically short traverses with a duration of 60 sec, run in replicate.  The short durations, bracketed by similar duration background measurements (when laser is not firing; aka gas blanks) makes it practical to model the predicted baseline variance when the standard replicates were actually measured. Thus, LODs are essentially based on the standard deviation of the baseline measurements bracketing the standard iterations, and then applied to model what baselines were like during the standard analysis. http://iolite-software.com/support/IoliteManualv3.09.pdf
* On GAM fit: I think it would be helpful to see an example of this fit.  We often do moving median followed by moving average smooths (using same boxcar length) to minimize effect of high-frequency outliers.  Reader would like to know how representative these fits are, so think an example would be beneficial to consider.
* How do we specifically identify biomineralization within the experiment period?  Do we only consider valve intervals since the last annuli was deposited?  Maybe I missed this, but we should document this. Are we comparing compositions for the entire valve interval (prismatic, nacre) as opposed to evaluating compositional variations within these intervals?  For example, if we see an increase in prismatic layer Pb toward the valve exterior, it might be muted if we present the average Pb for the entire interval as opposed to the last year of growth or last 100µm of growth.
* This section selectively documents if we could verify a change in composition (nacreous and/or prismatic layer) related to the experiment?  But, it also seems like there should be other foundational results presented, such as:
  * Results of specimen characterization: How old were they?  Annuli counting results
  *	How did specimen health vary over the experiment interval?  Was it aberrant?
  * How did specimen health relate to intra-watershed location?  This might be an important factor in interpreting results.
  *	How do elemental compositions vary for the same layers (regardless of the experiment).  How much within-layer variance; How does it change as a function of age? 
*	We have a major opportunity to document chemical composition variations associated with these Unionid bivalves – that will make a nice reference database for future studies.  Why not make this one of the project goals, in addition to the others regarding population decline?  By not tabulating this in the manuscript, we do not show the actual data very effectively (which is a challenge for such a large study).  This is also an opportunity to document the natural variability among specimens, which is an interpretation variable otherwise.
*	What about presenting graphical results showing detail for all transect for a few select individuals?  Similar to Fig 2.
  *	Seeing is believing and I think the reader might appreciate such a presentation.
* I’m finding it very difficult to understand the significance of A through D in the figure caption.  I think I understand what each figure generally shows with respect to the null hypothesis (although is D for prismatic layer during experiment only?).  It’s a lot of work to understand this – can we make it easier to follow for the reader?
* But we haven’t really evaluated the possibility that metal concentrations might change following biomineralization.  If outer shell regions, closest to ambient water, are most susceptible to such changes, the outermost portions of the prismatic calcite layer might have variations related to the change in environment associated with the experiment?  Another way to ask this questions might be do we see chemical gradients in the prismatic calcite layer?  If so, is it a natural result of the biomineralization process over time or could it be superimposed by the environment?
* In what?  Mn is a very redox sensitive element.  Changes in pH could affect Mn solubility.  Large runoff events (e.g., fertilizer) that increased nutrient contents in rivers might reduce dissolved oxygen and increase Mn solubility in the benthic realm?
* Again, we need to document specimen health early in the results section.  But. . . . . this potential makes it all the more important to evaluate the portions of the shell that may have experienced the effects of the ambient environment, even if they were not actively growing.

\doublespacing

```{r, echo = FALSE, message=FALSE}
library(elktoeChemistry)
library(gt)
```

# Introduction 

Freshwater mussels (Bivalvia: Unionidae, Magaritifera) are predominately suspension feeding bivalves that inhabit streams, rivers and lakes.Unionids provide a range of ecosystem services [@vaughn2017ecosystem] They serve as food for raccoons, muskrats and other species that inhabit riparian habitat, and when abundant help stabilize streambeds. While feeding unionids filter the water column removing algae, bacteria and detritus and process and sequester some contaminants [@gagne2007changes; @vaughn2017ecosystem]. Unionid shells support a microecosytem of bacteria, protozoa and fungi that play a prominent role in stream ecology. Their shells are similar to tree rings, archiving some aspects of their environmental history, and have been used as bioindicators of pollution [@markich2002freshwater; @brown2005freshwater; @shoults2014resolution; @wilson2018freshwater]

North American freshwater mussel species are imperiled throughout their home range. The majority, are federally or state listed as threatened, endangered or of special concern [@williams1993conservation]. Environmentally degrading riparian land-use practices have been attributed to their decline. Exposure to certain heavy metals from municipal effluents and other sources have been shown to increase morbidity and mortality in freshwater bivalves [@naimo1995a-review]. @hartmut2007declining attributed declining populations of the freshwater pearl mussel (*Margaritifera margaritifera*) to heavy metals and pesticides. Exposure to heavy metals have also been associated with genotoxic effects [@black1996dna; @khan2019bioaccumulation] enzymatic function, and behavioral changes [@naimo1995a-review].

The Appalachian Elktoe (*Alasmidonta raveneliana*) (Bivalvia: Unionidae) is native to the streams of eastern Tennessee and western North Carolina. The Elktoe is listed as federally endangered and remaining populations are found in the Nolichucky River in eastern Tennessee and the Upper Tennessee River system in North Carolina. Although once abundant in the Little Tennessee River, NC, precipitous declines were noted in  *A. raveneliana*  populations after major hurricane-related flooding associated with Hurricane Frances in 2004. However, within the same watershed’s  *A. raveneliana*  Tuckasegee River populations appeared stable. @miller2013concentrations evaluated sediment concentrations of Cu, Cr, Ni, and Zn in the Little Tennessee and Tuckasegee. When considered with the chemical concentrations of *A. raveneliana* shells measured by @jarvis2011water, they found these elements (Cu in particular) unlikely to be the primary culprit in the *A. raveneliana* decline. Reasons for the decline remain unclear [@jarvis2011water; @miller2013concentrations; miller2016potential; @pandolfi2016effects]. 

To determine how river-specific factors may be associated with health and physiological outcomes, we conducted a completely randomized experiment by assigning eighty-two  *A. raveneliana*  along with an equal number of hatchery-raised Wavy-Rayed Lampmussel (*Lampsilis fasciola*) to sites within each river [@salger2020]. In this paper, we ask whether exposure to the experimental sites modified the accretion of twenty trace elements in specimens’ valves. We demonstrate how chemical trajectories obtained by laser ablation inductively coupled plasma mass spectrometry (LA-ICP-MS) transects of valve cross-sections [@pozebon2014review] can be aligned with anatomical features of the valves in order to compare specific valve layers and time periods. 

# Materials and methods

## Study Specimens

Eighty-two *A. raveneliana* specimens were collected from a single location in the Tuckasegee river [@salger2020]. Ten were randomly selected and sacrificed for baseline comparisons, and the remaining 72 were randomly assigned and relocated in cages to six experimental sites, three each in the Tuckasegee and Little Tennessee Rivers (Figure \@ref(fig:studyMap)). The same number of young adult *L. faciola* specimens, obtained from the NC Wildlife Resources' Marian hatchery, were randomly assigned to a separate cages at the six sites. The plastic mesh cages were anchored in the bottom substrate. All animals were individually identified with randomly selected numeric identification tags glued to the shells. Weight and other metrics were obtained before and after deployment in the cages. After the experiment, a small (approximately 0.5 inch) section of the right valve from each animal was removed and embedded in plastic for laser ablation analyses.

```{r studyMap, echo = FALSE, fig.cap= "TODO: create a map of experimental sites."}
knitr::include_graphics(here::here("manuscript", "figures", "study_map.pdf"))
```

## LA-ICP-MS

Since the study was initiated to identify factors potentially contributing to the decline of the Elktoe populations, our analysis focused on the sections of shell that biomineralized during the period of the sentinel experiment. Parallel transects from the outer margins of valves, away from the umbo were targeted for analysis. For thicker valves we also analyzed additional offset scans to evaluate reproducibility and include older growth in the prismatic layers. In most cases, the entire preserved growth history was recorded by making multiple offset traverses (Figure \@ref(fig:alignment)). We used a 25$\mu$m round aperture moving at 5$\mu$m/sec.  The quadrupole sampling period (0.5762 s) enabled a moving compositional average for the 20 surveyed analytes every 2.88 $\mu$m (8.7 readings per 25$\mu$m footprint). For valve cross-sectional thicknesses on the order of 1 mm, transects included roughly 350 measurements. The laser trajectories were drawn perpendicular to annular growth rings in a "step-scanning" approach [@shoults2014resolution], moving from the inner to the outer shell, beginning and ending in epoxy. Scans varied depending on location, in terms of residing solely with the outer prismatic layer and periostracum, or being compound scans initiating in the nacreous layer. 

Each transect resulted in a chemical time-series for each analyte (concentration vs. distance; a matrix of about 7000 concentration measurements per mm). For data reduction^[Nathan/Heather - can you explain what this means?] we assumed a stoichiometric CaCO3 composition of 39.547 wt\% Ca in nacreous tissue and 40.0432 wt\% Ca in other tissue and did not adjust for mineralogy (aragonite vs. calcite){>>Include EPMA work done to evaluate Ca concentrations of the nacreous and prismatic layers.  The nacreous layer compositions would need a slight correction for the difference in Ca concentration (used as the internal standard).<<}. We used USGS MACS-3 (synthetic aragonite) as the primary calibration standard, and analyzed NIST 614 (trace elements at ~1 ppm) and NIST 612 (trace elements at ~35 ppm). For drift compensation, each hour of data collection on unknowns was bracketed by analyses of each standard (in duplicate). We surveyed the shells for twenty industrial metals and bio-metals: Mg, Ca, Cr, Mn, Fe, Co, Ni, Cu, Zn, As, Se, Kr (for Se interference correction if detected), Sr, Mo, Cd, Sn, Ba, Hg, Pb and U. Analyses were conducted at the University of Texas at Austin using an ESI New Wave Research NWR193 excimer laser coupled to an Agilent 7500ce ICP-MS. 

## Anatomical Alignment

Review of high resolution cross section images and annotation of the initial shell/epoxy boundary, annuli, nacre/prismatic layer boundary and terminal shell/epoxy boundary were made using Adobe Illustrator (Adobe Inc, San Jose CA). Annuli were identified as any linear growth feature which could be traced from its origin in the umbo to its emergence through the prismatic and periostracum layers. Non-continuous linear features along the transect were not annotated and assumed to be "false annuli" caused by environmental disturbances [@veinott1996identification]. 

To align anatomical features with chemistry data, several algorithms were used to first identify a reference point such as the epoxy/nacre boundary in the chemistry data. These methods included detecting changes in mean, variance [@killick2014changepoint; @killick2016changepoint] or the extrema [@borchers2018pracma] of Ca43_CPS, Pb208_CPS, or the ratio of the two (Figure \@ref(fig:alignment)). The difference of the distance between the detected reference point and the distance between the valves edges measured from high resolution scans was then computed. For each valve, the method that minimized this distance was then used to locate anatomical locations in the chemical data. The supplementary website includes the code and complete details of the alignment steps.

```{r alignment, echo = FALSE, fig.cap= "TODO"}
knitr::include_graphics(here::here("manuscript", "figures", "figure_alignment.pdf"))
```

## Chemical data pre-processing and de-noising

Valves were grouped into seven batches for LA-ICP-MS processing. During each batch process, the lower limit of detection (LOD) was assessed multiple times using the standards noted above. Concentration values were left-censored based on the mean LOD within a batch for each element. For analysis, parts-per-million (ppm) values were rescaled to $\text{mol/Ca mmol}$ where $\text{Ca mmol}$ was based on Ca\% of ~39.54\% for nacral observations and 40.078\% for all other layers.

To denoise the chemical signals, a Generalized Additive Model [@wood2017gam] was fit separately for each transect and each element with a tensor spline term for transect distance and term for anatomical layer. Fitted values from these models were used in subsequent analyses. 

## Study questions

We hypothesized that the distributions of one or more elements accrued in either the nacre or prismatic layer would differ between experimental sites. To examine differential accrual, we probed the research hypotheses by systematically examining both the sources and validity of any observed differences in shell composition in manner that was faithful to the completely randomized experimental design, thus permiting randomization-based inference [@splawa-neyman1990on-the-application; @coppock2019ri2] by permuting assignments to sites. 

Systematic differences in nacral elements accrued *prior* to the experiment could bias differences accrued during the experiment. Thus, to check that the sources of variation in the experiment were not due pre-experiment differences we first asked if there were differences in elemental concentrations *excluding* nacre accrued during the experiment. We examined this question both with (A) and without (B) baseline specimens which were not exposed to the experiment. Our primary focus was on whether nacre accrued during the experiment showed differences in elemental concentrations among and between experimental sites (C).

Laser transects did not always cross the prismatic layer in the region of most growth. However, we still tested the hypothesis of no difference between prismatic layer concentrations (D).

## Statistical Analyses

Null hypotheses of no difference between elemental distributions were tested using randomization-based inference using $m = 1000$ simulations from the experiment's distribution. Each analysis was conducted in the following series of steps: (1) the observed test statistic ($t_{obs}$) was computed using the observed assignments, (2) assignments were permuting by assigning specimens to different sites and test statistics re-computed, (3) repeat (2) $m$ times. The two-sided p-value was the proportion of times $t_{obs}$ was greater than or equal to the test statistic's distribution. The `ri2` R package was used to carry out inference [@coppock2019ri2; @rcore2019].

The test statistic (ri$_{ls}$) was an $F$-statistic comparing two least squares regressions, one including terms for experimental sites and  nested model which did not include terms. The dependent variables were each specimen's mean element concentration from observations in the region of interest. For example, for questions (Table \ref{test-questions}) involving the experimental period the youngest of all annuli was used; however, for questions regarding the pre-experiment time, the second youngest annuli was used. Weights in the regression were the sample variance taken over the same observations used to compute the mean. The models used included terms for a specimen's volume (height $\times$ width $\times$ length) measured at baseline and its LA-ICP-MS batch.

\begin{table}
\label{test-questions}
\caption{Data and test statistics used for randomization inference}
\begin{tabular}{lp{5cm}p{7cm}}
Label & Contrast & Transect Data Used  \\
\hline \\
A & Baseline vs. Sites & All nacre in baseline specimens and nacre excluding most recent growth in site specimens \\
B & Sites & nacre excluding most recent growth \\
C & Sites & only nacre of most recent growth \\
D & Sites & prismatic layer \\
\hline
\end{tabular}
\end{table}


The methodology, test statistics, and the model specifications are further described in the Supplementary Material.

# Results

Table XXX shows characteristics of the specimens at baseline.

```{r, results='asis', echo = FALSE}
val <- function(x){
  sprintf("%.f1 (%.f2)", mean(x), sd(x))
}

mussels_wide %>%
  mutate(
    weight_0 = if_else(species == "A. raveneliana", buoyant_weight_g_0, dry_weight_g_0)
  ) %>%
  group_by(species, site) %>%
  summarise(
    weight = val(weight_0),
    length = val(length_mm_0),
    width  = val(width_mm_0),
    height = val(width_mm_0),
    .groups = "keep"
  ) %>%
  gt() 
# %>%
#   tab_header(
#     title = "Specimen baseline characteristrics",
#   )
```


The experimental p-values corresponding to hypothesis tests for Q1-Q4 for each element are shown in Figure \@ref(fig:rires). Only *L. fasciola* had any results with p-values $<0.01$: Barium (B, C, D), Strontium (A, B), and Manganese (D). 

```{r rires, echo = FALSE, fig.cap = "Points are p-values (log$_{10}$ scale) testing the null hypothesis of no difference in each element accrued prior to the experiment in (A) nacre for sites versus baseline specimens excluding most recent annuli in non-baseline specimens, (B) nacre for between experiment sites excluding baseline excluding most recent annuli, (C) nacre between experiment sites excluding baseline including most recent annuli, and (D) prismatic layer between experiment sites. Smaller p-values suggest a larger discrepancy between sites under a null hypothesis of no difference between sites. P-values $<0.1$ are labelled by element."}
knitr::include_graphics(here::here("manuscript", "figures", "ri_ncr_hypothesis_v002.pdf"))
```

Figure \@ref(fig:risummary) shows mean values from each specimen in the regions of interest for questions C and D where at least one p-value was $<0.01$ for either species. Figure \@ref(fig:risummary) reveals that the differences in *L. fasciola* Ba concentrations are due to low concentrations among specimens from the first Little Tennessee site in both the nacre and prismatic layer. Similarly, differences in *L. fasciola* Mn concentration appear to be due to higher prismatic layer concentrations at the first Little Tennessee site. 

```{r risummary, echo = FALSE, fig.cap = "Open points are mean concentrations in a valve region of interest for each specimen. The diamond is the mean of the specimen means within a site."}
knitr::include_graphics(here::here("manuscript", "figures", "ri_ncr_summary_v002.pdf"))
```

# Discussion

Based on current understanding of bivalve shell growth [@checa2000a-new-model], the laser transects that crossed the nacre should have captured the accretion that occurred during the experiment. By careful alignment of anatomic imaging and chemical data obtained by LA-ICP-MS, we were able to extract and analyze elemental concentrations from the nacre, including the locations of annuli within the nacre. We hypothesized that some elements would differentially accrue within the nacre across experimental assignments and could be detected from LA-ICP-MS data. @jackson2003laser, for example, experimentally fed contaminated fish to banded water snakes and showed LA-ICP-MS used on tail tissue was able to differentiate between treatments.

Our analysis revealed no clear cut differences between sites that could be easily explained by the experiment itself. The lower concentrations of Barium in Little Tennessee Site 1 *L. fasciola* specimen has no obvious explanation, for example. All *L. fasciola* specimens were hatchery raised in the same conditions. The shells for the experimental sites were randomly distributed in the LA-ICP-MS sequence, suggesting the result is not an artifact of LA-ICP-MS processing. Moreover, *A. raveneliana* specimens do show a similar pattern in the nacre. Since *A. raveneliana* were much older than *L. fasciola*, comparing prismatic layer difference across species may not be relevant, as the laser transect in general crossed older prismatic layer sections in *A. raveneliana*. 

We were particularly interested in Mn, as it is an essential trace element with important roles in numerous enzymatic pathways that could have harmful effects on freshwater bivalves. Interestingly, surviving valves from the Little Tennessee had films on the gills that were apparent by scanning electron microscopy, which upon x-ray diffraction analysis consisted largely of Mn [@levine2020].

Manganese valve concentrations have been shown to correlate with seasonal changes [@siegele2001manganese; @shoults2014resolution]. The reasons for the seasonal variation are unknown, but in our study, different survival times could have resulted in different accrual patterns. Those that survived longer would have kept accumulating Mn for later in the season. We do not know when specimens in the experiment died. It could be that those specimens that survived to the end of the study (Tuckasegee 1 site had the lowest mortality) were accruing shell for longer while the trajectories for other specimens were censored at their time of death.

We assumed that the youngest annuli in the nacre accrued during the experiment, but the data provide no way of testing this assumption. In Little Tennessee specimens, where high morbidity and mortality was observed in *A. raveneliana* [@salger2020], it could be that little to no nacre accrued during the experiment. Marking the start of the experiment shells in some way may have provided appropriate data, but @haag2008testing points out that notching shells does harm individuals.

* wrap up

# Funding

This work was supported by funds provided by the North Carolina Department of Transportation [NCDOT Project 2013-2012-37]; and the US Fish and Wildlife Service [F18AC00237].

# Acknowledgment

The authors thank [TODO]


# References