---
title: "Data Preparation"
author: "Bradley Saul"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params: 
   file_baseline_specimens: "../extdata/Mussel Metrics Data - Baseline.xlsx"
   dir_laicpms_source: "../extdata/Data/Geochemical Time-Series"
   file_out_chemistry: "../data/valve_chemistry.rds"
   file_out_lod: "../data/lower_detection_limits.rds"
   file_lod: "../extdata/Std recoveries comp D1-7 reproc.xlsx"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This document details the process of preparing the valve data for analysis.

# Collecting Necessary Data

The `elktoeChemistry` R package contains the source data and functions needed for the data processing.

```{r message=FALSE}
library(elktoeChemistry)
library(stringr)
```


## Covariates and Outcomes

The `elktoeChemistry` packages includes a `data.frame` with information on the study specimens exposed to the experiment. Variables ending in:

* `_0` are measurements at baseline;
* `_1` are measurements when the specimen was pulled from experiment;
* `_d` are differences between baseline and end (`*_1 - *_0`).

```{r}
str(mussels_wide)
```
  

### Add baseline specimens

The `mussels_wide` dataset does not include data on those specimen held out at baseline. The file ``r params$file_baseline_specimens`` contains these data. We add these data to `mussels_wide`.

```{r}
baseline_specimens <- 
  readxl::read_xlsx(
    path = params$file_baseline_specimens,
    skip = 1L,
    col_names = c("id", "weight_g_0", 
                  "length_mm_0", "width_mm_0", "height_mm_0", 
                  "gravid", "species")
  ) %>%
  filter(!is.na(id)) %>%
  mutate_at(2:5, as.numeric) %>%
  mutate(
    buoyant_weight_g_0 = if_else(species == "A. rav.",
                                 weight_g_0,
                                 NA_real_),
    dry_weight_g_0 = if_else(species == "L. fas.",
                             weight_g_0,
                             NA_real_),
    volume_0 = length_mm_0 * width_mm_0 * height_mm_0,
    species = if_else(species == "A. rav.",
                      "A. raveneliana",
                      "L. fasciola")
  ) %>%
  select(-weight_g_0, -gravid)

# Append to mussels_wide
mussels_wide <-
  mussels_wide %>%
  bind_rows(baseline_specimens)
```

## LA-ICP-MS Chemistry data

The chemical time series data from the LA-ICP-MS runs are contained ``r params$dir_laicpms_source``. Each transect is contained in a separate `xlsx` file. Some transects are excluded from analysis due to the shape of the transect or other irregularities seen in the scans.

```{r}
# Files to exclude from analyses
exclude_files <- c(
   # Horizontal (fly-scanning) transects
   "2-A4-4", "8-C474-5", "27-C531-2", 
   "11-C482-1", "11-C483-1",
   
   # Excluding C497 due to unusual structure at the mantle edge
   "16-C497-1", "16-C497-2", "16-C497-3",
   
   # ??? Per notes: This shell does not exist.
   "45-P139-2" 
)
```


```{r}
valve_files <- list.files(
  path         = params$dir_laicpms_source,
  full.names   = TRUE,
  recursive    = TRUE,
  include.dirs = FALSE)

# remove exclusions
valve_files <- valve_files[!str_detect(valve_files, glue::glue_collapse(exclude_files, sep = "|"))]
raw_data    <- purrr::map(valve_files, readxl::read_excel)
sheets      <- purrr::map_chr(valve_files, readxl::excel_sheets)
file_names  <- str_extract(valve_files, "(?<=[1-9]/).*(?=\\.xlsx)")
names(raw_data) <- file_names
```

### Preimport consistency checks

```{r}
# Check sheet names == file name
if(length(sheets) !=  length(file_names)){
  warning("Different # of files and sheets")
}

valve_files[!str_detect(file_names, sheets)]
```

```{r}
# Check that the column names are consistent
colNames <- purrr::map(raw_data, names)
index_nomatch <- !purrr::map_lgl(colNames, ~ all(.x %in% colNames[[1]]))

if(any(index_nomatch)){
  warning("Some columns don't match")
}

colNames[index_nomatch]
```

### Import chemistry data

```{r}
valve_chemistry <- 
  bind_rows(raw_data, .id = "file_name") %>%
  dplyr::select(
    file_name, 
    time = ElapsedTime_s, 
    scan_distance =`Dist (µm)`,
    note = Notes, 
    everything()
  ) 

str(valve_chemistry)
```

### Post-import consistency checks

```{r}
hold <- valve_chemistry %>% filter(!is.na(scan_distance)) %>%
  group_by(file_name) %>%
  mutate(
    x_time = c(0.576, diff(time)),
    x_dist = c(2.88, diff(scan_distance)),
    has_on = str_detect(note, "[Oo]n"),
    has_off = str_detect(note, "[Oo]ff")
  )

# Are all the elapsed times the same within 0.01
hold %>%
  summarise(
    same_time_diffs = all((x_time - 0.01) < x_time[2] & 
                           x_time[2] < (x_time + 0.01), na.rm = TRUE)
  ) %>%
  filter(!same_time_diffs) %>%
  nrow() %>%
  {if(. > 0) warning("not all the elapsed time are within 0.01")}

# Are all the scan distances within 0.01?
hold %>%
  summarise(
    same_dist_diffs = all((x_dist - 0.01) < x_dist[1] & 
                           x_dist[1] < (x_dist + 0.01), na.rm = TRUE)
  ) %>%
  filter(!same_dist_diffs) %>%
  nrow() %>%
  {if(. > 0) warning("not all the scan distances are within 0.01")}

## Do all files have an "on" and "off" indicator in the note field?
hold %>%
  summarise(
    n_on    = sum(has_on, na.rm = TRUE),
    n_off   = sum(has_off, na.rm = TRUE),
    any_on  = any(has_on, na.rm = TRUE),
    any_off = any(has_off, na.rm = TRUE)
  ) %>%
  filter(
    !any_on | !any_off | n_on > 1 | n_off > 1
  ) %>%
  nrow() %>%
  {if(. > 0) warning("not all files have on and off indicators")}

```

### Prepare chemistry data for merging with valve measurements

Next, we identify the laser on/off points which were manually annotated in the `xlsx` files and prepare the `valve_chemistry` data for further processing. The result is written to ``r params$file_out_chemistry``

```{r}
valve_chemistry <-
  valve_chemistry %>%
  group_by(file_name) %>%
  # Identify laser on/off points
  mutate(
    is_laser_on  = str_detect(note, "[Oo]n$"),
    is_laser_off = str_detect(note, "[Oo]ff$"),
    is_laser_on  = if_else(is.na(is_laser_on), FALSE, is_laser_on),
    is_laser_off = if_else(is.na(is_laser_off), FALSE, is_laser_off),
    on_row       = which(is_laser_on),
    off_row      = which(is_laser_off),
    last_val_row = min(which(is.na(Ca43_CPS)) - 1),
    rn           = row_number(),
    # scan_distance is NA before the on_row, add it back
    scan_distance  = if_else(
      is.na(scan_distance), 
      -2.881 * (on_row - rn) ,
      scan_distance)
  ) %>%
  # Remove rows at end of each file with no data
  filter(rn <= last_val_row) %>%
  dplyr::select(
    -is_laser_on, -is_laser_off, -rn, -on_row, -off_row, -last_val_row
  ) %>%
  ungroup() %>%
  mutate(file_name = clean_ids(file_name)) %>%
  tidyr::separate(file_name, sep = "-", into = c("id", "transect")) %>%
  dplyr::select(id, transect, distance = scan_distance, everything(), -time, -note)

saveRDS(valve_chemistry, file = params$file_out_chemistry)
```

## Lower limit of detection data

The excel file ``r params$file_lod``  was prepared by the UT LA-ICP-MS lab and contains lower limit of detection information data. Here, we use his data to find the average per element per 
drawer and write the results to ``r params$file_out_lod``.

```{r}
readxl::read_excel(
   path = params$file_lod,
   sheet = "Without Stats"
  ) %>%
  mutate(
    standard = stringr::str_extract(Comments, "614|612")
  ) %>%
  group_by(standard, Date) %>%
  mutate(
    standard_run = 1:n(),
  ) %>%
  ungroup() %>%
  select(-Date, -Time, -"Duration(s)", -DateTime, -Comments) %>%
  filter(Exclude == 0) %>%
  select(
    drawer = Drawer, output = Output, source_file = `Source file`, 
    standard, standard_run, matches("CPS|ppm")
  )  %>%
  tidyr::gather(
    key = "key", value = "value", 
    -drawer, -output, -source_file, -standard, -standard_run,
     na.rm = TRUE) %>%
  mutate(
    drawer       = as.numeric(stringr::str_remove(drawer, "D")),
    measure      = stringr::str_replace(stringr::str_extract(key, "_(Int2SE|LOD)$"), "_", ""),
    measure      = if_else(is.na(measure), "raw", measure),
    element      = stringr::str_replace(key, "_(Int2SE|LOD)$", "")
  ) %>%
  select(
    drawer, standard, output, source_file, standard_run, element, measure, value
  ) %>%
  # Just keep the lower limit of detection
  filter(measure == "LOD") %>%
  group_by(element, drawer, standard) %>%
  summarise(
    lod_mean = mean(value)
  ) %>%
  group_by(element, drawer) %>%
  summarise(
    lod = mean(lod_mean)
  ) %>%
  saveRDS(file = params$file_out_lod)

```
