---
title: "Ad Hoc Analysis of Net Intensity Data"
author: "Bradley Saul"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Net Intensity}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: vignettes.bib
editor_options: 
  chunk_output_type: console
---

This document does a preliminary, ad-hoc analysis of the data in
`"extdata/Data/Mussel Shell Edge Net Intensities.xlsx"`,
which contains:

* net intensity values (not concentration);
* these values have been summarized by mean and standard deviation
within nacre and periostracum 
within valve/transect;
* the nacre and periostracum layers were *not* identified 
using the automatic process described in the alignment document.

```{r, message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(grid)
library(gridExtra)
library(gt)
```

## Load data

```{r load-data, message=FALSE}
xldt <- readxl::read_excel(
  path = here::here("extdata/Data/Mussel Shell Edge Net Intensities.xlsx")
)
```


*TODO*: resolve questions about data

* The `sample` column has inconsistencies.
I (BS) understand the common pattern to be "`{valve_id}.{transect_id}{n|p}`.
But, for example, `A3.1na`, `A5.1p_1` and `P111.1P` do not conform to this pattern.
* I ignored these inconsistences for now, 
and assume they are meant to conform to the common pattern.

Bring in the `valve_data`, which contains information on sites and outcomes.

```{r}
valve_data <- 
  readRDS(here::here("data/valve_data.rds")) %>%
  select(id, species, river, site, site_num, dead, moribund) %>%
  distinct()
```

## Prepare data for analysis

```{r}
# Extract the mean value and reshape the data to make it easier to analyze.
dt <- 
  xldt %>%
  select(
      drawer = Drawer
    , sample = Sample
    , duration = `Duration (s)`
    , width = starts_with("Width")
    , ends_with("_mean")
  ) %>%
  # HARDCODES
  mutate(
    sample = case_when(
        sample == "P111.1P" ~ "P111.1p"
      , TRUE ~ sample
    )
  ) %>%
  filter(
    # Don't know what layer this is
    !(sample %in% c("P164.2")) 
  ) %>%
  mutate(
      id = stringr::str_extract(sample, "^[A|C|P]\\d{1,3}")
    , transect = stringr::str_extract(sample, "(?<=\\.)[\\d]")
    , layer = tolower(stringr::str_extract(sample, "[n|p]"))
  ) %>%
  pivot_longer(
      cols = ends_with("_mean")
    , names_to = "element"
    , values_to = "value"
  ) %>%
  mutate(
    element = stringr::str_replace(element, "_CPS_mean", "")
  ) %>%
  left_join(valve_data, by = "id")
```

## Test for differences in net intensity per element

* Remove baseline specimens
* Average intensities (mean) across transects within valves
* Remove `TotalBeam`

```{r}
tdt <-
  dt %>%
  filter(site != "Baseline", element != "TotalBeam") %>%
  group_by(species, river, site, element, id, layer) %>%
    summarise(
      # TODO: weight by duration (?)
      value = mean(value),
      .groups = "drop"
  )  %>%
  group_by(species, element, layer)

make_pvalue_table <- function(x){
  x %>%
   tidyr::pivot_wider(
    names_from = c(species, layer),
    values_from = p
  ) %>%
  ungroup() %>%
  gt() %>%
  cols_label(
    `A. raveneliana_n` = "nacre",
    `A. raveneliana_p` = "perio",
    `L. fasciola_n` = "nacre",
    `L. fasciola_p` = "perio"
  ) %>%
  fmt_number(
    columns = matches("_p|_n"),
    decimals = 4,
  ) %>%
  tab_spanner(
    label = "A. raveneliana",
    columns = c("A. raveneliana_n", "A. raveneliana_p")
  ) %>%
  tab_spanner(
    label = "L. fasciola",
    columns = c("L. fasciola_n", "L. fasciola_p")
  )
}
```

### among all sites, within species/layer

The values in the table are p-values from a 
[Kruskal-Wallis test](https://en.wikipedia.org/wiki/Kruskal%E2%80%93Wallis_one-way_analysis_of_variance),
a rank-based ANOVA-like test.

```{r, message = FALSE}
tdt %>%
  summarise(
    p = kruskal.test(value ~ site)$p.value
  , .groups = "drop"
  ) %>%
  make_pvalue_table()
```

### between Tuck 1 and other sites, within species/layer

The values in the table are p-values from a 
[Kruskal-Wallis test](https://en.wikipedia.org/wiki/Kruskal%E2%80%93Wallis_one-way_analysis_of_variance),
a rank-based ANOVA-like test.

```{r}
tdt %>%
  summarise(
    p = kruskal.test(value ~ (site == "Tuck 1"))$p.value
  , .groups = "drop"
  ) %>%
  make_pvalue_table()
```

## Plots of intensities per element


* lines connect `median` values of each site.

```{r element_intensities}
make_element_plot <- function(x, element = NULL){
  
  if (is.null(element)) {
    x <- x
  } else {
    x <- x %>% filter(element == !! element)
  }
  
  ggplot(
    data = x, 
    aes(x = site, y = value, group = river)
  ) +
   geom_point(size = 0.25) +
   geom_line(
     data = x  %>% 
      group_by(species, river, site, layer, element) %>%
      summarise(
        value = median(value),
        .groups = "drop"),

     size = 0.1) 
}
```


```{r element_plot_data}
element_plot_data <-
  dt %>%
  group_by(species, river, site, element, id, layer) %>%
      summarise(
        # TODO: weight by duration (?)
        # Take mean across transects within a valve
        value = mean(value),
        .groups = "drop"
  )  %>%
  filter( site != "Baseline" ) %>%
  mutate(
    site = factor(
      site, 
      levels = c("Tuck 1", "Tuck 2", "Tuck 3", 
                 "LiTN 1", "LiTN 2", "LiTN 3"))
  )
```

```{r}
make_single_element_plot <- function(element){
  f <- function(species, element){
  element_plot_data %>%
  filter(species == !! species) %>%
  make_element_plot(element = element) +
    facet_grid(layer ~ ., scales = "free") +
    ggtitle(paste(species, "-", element)) + 
    theme(
      axis.text.y = element_blank(),
      axis.text.x = element_text(angle = 0),
      strip.text.y = element_text(angle = 0),
      axis.title.x = element_blank(), 
      axis.title.y = element_blank()
    )
  }
  grid.arrange(
    f("A. raveneliana", element),
    f("L. fasciola", element),
    nrow = 1
  )
}
```


```{r, results='asis', fig.width=8, fig.height=4}
elements <- dt %>%
  distinct(element) %>% 
  filter(element != "TotalBeam") %>%
  pull() %>%
  sort()

for(element in elements){
  
  cat("\n")
  cat("###", element, "\n")
  
  make_single_element_plot(element)
  cat("\n")

}
```

## Principle Components


```{r}
do_princ <- function(dt) {

  fdt <-
    dt %>%
    distinct(id, site, river)
  
  dt %>%
    select(id, element, value) %>%
    pivot_wider(
        id_cols = "id"
      , names_from = "element"
      , values_from = "value")  %>%
    .[, -1] %>%
    as.matrix() %>%
    prcomp(center = TRUE, scale = TRUE) -> 
  pr
  
  temp   <- hclust(dist(pr$x), method = "ward.D")
  mojema <- mean(temp$height) + 3.75 * sd(temp$height)
  g      <- length(temp$height[temp$height>mojema]) + 1
  if(g == 1){
    return(1)
  }
  
  list(  dt = fdt %>% mutate(PC1 = pr$x[, 1], PC2 = pr$x[, 2])
       , clust = cutree(temp, k=g))
}
```

### Periostracum in A. rav, excluding Baseline specimens

```{r, fig.height=4, fig.width=5}
element_plot_data %>%
  ungroup() %>%
  filter(
      layer == "p"
    , element != "TotalBeam"
    , species == "A. raveneliana"
    , site != "Baseline"
  )  %>%
  do_princ() %>%
  .[["dt"]] %>%
  ggplot(
    aes(x = PC1, y = PC2, color = site)
  ) + geom_point()
```

### Nacre in A. rav, excluding Baseline specimens

```{r, fig.height=4, fig.width=5}
element_plot_data %>%
  ungroup() %>%
  filter(
      layer == "n"
    , element != "TotalBeam"
    , species == "A. raveneliana"
    , site != "Baseline"
  )  %>%
  do_princ() %>%
  .[["dt"]] %>%
  ggplot(
    aes(x = PC1, y = PC2, color = site)
  ) + geom_point()
```




<!-- ```{r} -->
<!-- library(mediation) -->

<!-- meddt <-  -->
<!--   dt %>% -->
<!--   filter(site != "Baseline", element != "TotalBeam") %>% -->
<!--   group_by(species, river, site, dead, moribund, element, id, layer) %>% -->
<!--     summarise( -->
<!--       # TODO: weight by duration (?) -->
<!--       value = mean(value), -->
<!--       .groups = "drop" -->
<!--   ) %>% -->
<!--   mutate( -->
<!--     dead = if_else(is.na(dead), 1, dead) -->
<!--   ) -->

<!-- do_mediation <- function(dt, .trtv = "LiTN 3"){ -->
<!--   mfit <- lm(value ~ site, data = dt) -->
<!--   ofit <- glm(dead ~ value + site, data = dt, family = binomial()) -->

<!--   mediate( -->
<!--    model.m = mfit, -->
<!--    model.y = ofit, -->
<!--    treat = "site", -->
<!--    mediator = "value", -->
<!--    control.value = "Tuck 1",  -->
<!--    treat.value =  .trtv,  -->
<!--    robustSE = TRUE, -->
<!--    sims = 100) -->


<!-- } -->
<!-- ``` -->

<!-- ```{r} -->
<!-- medres <-  -->
<!-- meddt %>% -->
<!--   filter(species == "A. raveneliana") %>% -->
<!--   group_by( -->
<!--     species, element, layer -->
<!--   ) %>% -->
<!--   tidyr::nest() %>% -->
<!--   mutate( -->
<!--     medi = purrr::map(data, .f = ~ do_mediation(dt = .x) %>% summary() %>% tidy()) -->
<!--   ) -->

<!-- medres %>% -->
<!--   ungroup() %>% -->
<!--   dplyr::select(element, layer, medi) %>% -->
<!--   tidyr::unnest(cols = c(medi)) %>% -->
<!--   filter(term %in% c("acme_0", "acme_1")) %>%View() -->

<!--   # meddt %>% -->
<!--   # filter(species == "A. raveneliana", layer == "p", element == "Pb208") %>% -->
<!--   # do_mediation() %>% -->
<!--   # summary() %>% -->
<!--   # tidy() -->


<!-- ``` -->


