---
title: "Examine Censoring"
author: "Bradley Saul"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Element Summaries}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: vignettes.bib
params: 
   inputs: 
     value: 
       analysis: "../data/analysis_data.rds"
editor_options: 
  chunk_output_type: console
---

In this document, we examine the proportion of observations censored
(below the lower limit of detection)
within each valve's layer.
Elements with a high degree of censoring across all layers
may be removed from further analysis.


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo    = FALSE,
  message = FALSE
)
```


```{r}
library(elktoeChemistry)
library(dplyr)
library(ggplot2)
analysis_data  <- readRDS(params$inputs$analysis)
```

Use the `analysis_data` function to observations for

* all signal prefiltering methods (`signal = "all"`)
* all specimens (`contrast = "all"`)
* all available layers
* group observations by layer across transects within a specimen
(`group_by_valve = TRUE`)
* require at least 5 observations within a layer
(`.min_n_obs`)

```{r}
dt <-
  analysis_data(
    contrast = "all",
    test_data_FUN = function(data) data, # just return the data unmodified
    elements  = "all", 
    signals   = "all",
    group_by_valve = TRUE,
    transect_opts  = list(
      .layers    = c("ipx", "ncr", "psm", "pio", "opx"),
      .min_n_obs = 5L
    )
  ) 
```

```{r functions}
# Identify nacre wiithin the *yountest* annuli as "ncr_new"
# and nacre in older annuli as "ncr_old"
layer_modifier <- function(layer, annuli){
  case_when(
      layer == "ncr" & annuli == "A" ~ "ncr_new",
      layer == "ncr" & annuli != "A" ~ "ncr_old",
      TRUE ~ layer
  ) %>%
  factor(
    levels = c("ipx", "ncr_new", "ncr_old", "psm", "pio", "opx"),
    ordered =  TRUE)
}
```

Use this package's `summarize_specimens` function to compute
summary statistics by layer within each specimen.

```{r summarize}
specimen_summaries_by_signal <- 
  dt %>%
  group_nest(signal) %>%
  mutate(
    data = purrr::map(
      .x = data, 
      .f = ~summarize_specimens(.x, modify_layer = layer_modifier))
  ) %>%
  tidyr::unnest(cols = c("data"))
```

Examine the average `pc` (proportion censored) for each element by layer.

```{r}
avg_pc_by_signal <- 
  specimen_summaries_by_signal %>%
  group_by(species, signal, element, layer) %>%
  summarize(
    pc = mean(pc)
  ) 
```


```{r}
ggplot(
  data = avg_pc_by_signal,
  aes(x = signal, y = pc)
) +
    geom_point(size = 0.5, shape = 1) +
  facet_grid(element ~ layer) +
  theme(
    strip.text.y = element_text(color = "blue", angle = 0)
  )

```

GAM no good.

```{r plot}

specimen_summaries_by_signal %>%
  filter(signal == "base", species == "Arav") %>%
ggplot(
  aes(x = site, y = pc)
) + 
  geom_point(size = 0.5, shape = 1) +
  facet_grid(element ~ layer) +
  theme(
    strip.text.y = element_text(color = "blue", angle = 0),
    axis.text.x = element_text(angle = 90)
  )
```



